---
output: 
html_document:
  css: style.css
citation_package: natbib
keep_tex: false
fig_caption: true
latex_engine: pdflatex
title: "Short-term forecasts of cases and deaths of COVID-19 in multiple countries"
author:
- name: Sangeeta Bhatia and Pierre Nouvellet
affiliation: Imperial College London
abstract: 
keywords: 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: double
bibliography: 
biblio-style: apsr
endnote: no
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(drake)
library(kableExtra)
library(DT)
source("R/util.R")
raw_data <- drake::readd(raw_data)
raw_data <- dplyr::rename(raw_data, country = "Countries.and.territories")


```
# Introduction {.tabset .tabset-fade .tabset-pills}

As of `r format(Sys.time(), '%B %d, %Y')`, there have more than
351,000 cases of COVID-19 across the world, with more than 15,000
deaths. In this report, we use the reported number of deaths
of COVID-19 to make short term forecasts of the trajectory of the
outbreak.

## Summary

### Objectives and caveats

The **main** objective in this report is to produce forecasts of the number of deaths in the week ahead per country with active transmission.

*	Active transmission is defined as have more than five deaths in the last week and preceding week.
*	We forecast deaths as it likely that reporting of deaths in more reliable than reporting of cases, and temporal trends in reporting of deaths are also likely to be much smaller than temporal trends in reporting cases.
*Because we are forecasting deaths, latest estimates of transmissibility reflect the epidemiological situation when those dying were infected.

A **secondary** objective of this report is to analyse case ascertainment per country. In this analysis, key assumption are:
  
*	The true underlying Case Fatality Ratio (CFR) is 1%,
*	The delay from a death to this case being reported is 10 days.

### Key results on transmissibility and forecasting

Key results below are based on ensemble forecast.

Transmissibility is characterised by the reproduction number, i.e. the average number of cases that one infected is likely to infect. Analysis of transmissibility indicates that the reproduction number last week (week starting [date]) was highest in:
  
*	[Top country name] with an estimated Rt of: [median; 95%CrI [] ], and
*	[2nd Top country name] with an estimated Rt of: [median; 95%CrI [] ],

and was lowest in:
  
*	[Bottom country name] with an estimated Rt of: [median; 95%CrI [] ], and
*	[2nd Bottom country name] with an estimated Rt of: [median; 95%CrI [] ]

Forecast of predicted deaths in the coming week (week starting [date]) is expected to be highest in:
  
*	[Top country name] with a forecasted number of weekly deaths Rt of: [median; 95%CrI [] ], and
*	[2nd Top country name] with a forecasted number of weekly deaths Rt of: [median; 95%CrI [] ],

and was lowest in:
  
*	[Bottom country name] with a forecasted number of weekly deaths Rt of: [median; 95%CrI [] ], and
*	[2nd Bottom country name] with a forecasted number of weekly deaths Rt of: [median; 95%CrI [] ],

Forecasts in previous weeks were good with [XX]% of observed numbers of deaths per country included in our 95% CrI.

### Key results on case ascertainment

Case ascertainment based on the last previous 2 weeks deaths and reported cases in the 10 days prior to that period indicate that countries with lowest case ascertainment were:
  
*	[Top country name] with an estimated ascertainment of: [median%; 95%CrI []% ], and
*	[2nd Top country name] with an estimated ascertainment of: [median%; 95%CrI []% ],

and countries with the highest case ascertainment were:
  
*	[Top country name] with an estimated ascertainment of: [median%; 95%CrI []% ], and
*	[2nd Top country name] with an estimated ascertainment of: [median%; 95%CrI []% ],

Based on the estimated ascertainment, we estimated the true size of the epidemics in the previous 7 days (week starting [date]). 
Countries with the highest true epidemics size in this period were:
  
*	[Top country name] with an estimated true epidemics size in this period of: [median; 95%CrI [] ], and
*	[2nd Top country name] with an estimated ascertainment of: [median; 95%CrI []%],

and countries with the lowest true epidemics size in this period were:
  
*	[Bottom country name] with an estimated true epidemics size in this period of: [median; 95%CrI [] ], and
*	[2nd Bottom country name] with an estimated ascertainment of: [median; 95%CrI []%],


## Methods 

The objective of this project is to forecast for the daily number of COVID-19 deaths in countries with sustained local transmission. We aim to produce weekly forecasts every Monday. Multiple teams are involved, each producing weekly forecasts and a centralised team also produces an ensemble forecast.

We use reported deaths rather than pretored cases, as the former ascertainment of the former is likely to be much more stable over time than the later. The ascertainment of deaths is also likely to be much higher then for cases, as most cases will show mild (if any) symptoms.

As we forecasting deaths, the latest estimates of trasnsmissibility reflect the epidemiological conditions when those dying were infected. Therefore, adisadvantage in using death, is this delay between case reporting and death.

As weel as forecasting, we also use reported death and past report cases (using a fix delay) to analyse the reporting trends per country. If the reporting of cases and death was perfect, as well as the delay between reporting and death fixed and known, the ratio of death to delayed cases would be exactly the Case Fatalty Ratio (CFR).

Forecasts are produced weekly for based on data available on the last Sunday of the previous week. Forecasts are produced only for countries that reported at least 5 deaths in the previous week and 10 deaths in the previous two weeks. For the purpose of our analysis, we assume those countries show strong evidence active and sustained local transmission.

Importantly, all weekly forecasts rely on estimation of current level of transmissibility. Given we are forecasting number of daily deaths expected, current transmissibility reflects epidemiological dynamics when the cases who died became infected. 

Therefore, current estimates of transmissibility based on death reflect transmission events that occur about two to three weeks ago, given the delays between infection and death. As a consequence, in the best case scenario, the impact of current control on estimated of transmissibility will be quantifiable with a two to three weeks delay.

In the results section of each model output, a 'methods' section outlines briefly the methodology used to produce forecasts.



While death counts are used for the main analysis and producing forecasts. Reported cases and the death to reported cases ratios for each country are show in the 'Supplementary Information' section.


## Projections and Rt Estimates {.tabset .tabset-fade .tabset-pills} 

### Ensemble Model {.tabset .tabset-fade .tabset-pills} 

#### Projections

Below for each country with active transmission (see methods), we plot the observed incidence of deaths in the last 3 weeks (black dots). Past forecasts, where available, are shown in green (median and 95%CrI band), while latest forecasts are show in red. Horizontal dashed line show the start and end of each week (Monday to Sunday).

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}

obs <- drake::readd("obs")
ensb_pred <- drake::readd("ensemble_model_qntls")
ensb_pred1 <- ensb_pred[ensb_pred$si == "si_1", ]
ensb_pred1$date <- as.Date(ensb_pred1$date)
projection_plot(obs, ensb_pred1)

```

#### Summary of results

Below for each country with active transmission (see methods), a table summarising for each period and country forecasted, the observed (where available) and forecasted weekly death count, and estimated level of transmissibility.

```{r echo = FALSE, warning = FALSE, message = FALSE}
x <- drake::readd("fmtd_ensemble_weekly_qntls")[[1]]

## knitr::kable(
##     x,
##     align = "l",
##     digits = 2
## ) %>% kable_styling()

DT::datatable(x)

```

### Model 1 {.tabset .tabset-fade .tabset-pills} 

#### Methods


**Estimating current transmissibility**

We use a previously described approach [cori et al. 2013] to quantify transmissibility during the calibration period, from the incidence of death time series, assuming a certain distribution for the serial interval for death (the time between death in a case and death in their infector). Here, we assumed that the serial interval distribution for death was similar to the classical serial interval distribution (the time between onset of symptom in a case and onset of symptom in their infector).

Our main analysis assume a gamma distributed serial of mean 4.8 days and standard deviation of 2.7 days following [Wang and Teunis, 2020].

**Include figure of histogram of SI**

The serial interval above might be biased toward lower values given the observation bias whereby, in contact tracing studies, long serial interval tend to be under-represented. To account for this, as a sensitivity analysis, we also ran our analysis with an alternative longer serial interval of mean 6.48 days and standard deviation of 3.83 days [Ferguson 2020, report 9]. Results for this longer interval are presented in the supplementary information section. Most importantly, while using such longer serial interval has very little impact on the weekly forecast produced, it results in much higher estimated of transmissibility.

We assumed that transmissibility was constant throughout the calibration period, and measured it through the reproduction number, R, defined as the average number of secondary cases infected by an infected individual. The estimate of R is informative as if R is above the threshold value 1, and remains above 1, the outbreak is likely to grow further, whereas if R is below 1, and remains below 1, the outbreak will die out.

Our method assumes that the daily incidence can be approximated by a Poisson process using the so-called renewal equation:

$$ I_t \sim Pois\left( R_t \Sigma_{s=1}^{t} I_{t-s} w_s \right) $$


where $I_t$ is the incidence (of deaths) on day $t$, $R_t$ is the reproduction number on day $t$, and $w$ is the probability mass function of the serial interval. 

**Forward projections**

For each country, we used the renewal equation to project the incidence forward, given an estimated reproduction number, and the observed incidence of death. We sampled a set of reproduction numbers from the posterior distribution obtained in the estimation process. For each of these sets, we simulated stochastic realisations of the renewal equation from the end of the calibration period; leading projected incidence trajectories. 

Projections were made on a 7-days horizon. The projections assume that the transmissibility remains constant over this 7-days horizon. If transmissibility were to decrease as a result of control interventions and/or changes in behaviour over this time period, we would predict a lower number of cases; similarly, if transmissibility were to increase over this time period, we would predict a higher number of cases. We limited our projection to 7 days only as assuming constant transmissibility over longer time horizons seemed unrealistic. 

Super-spreading has been shown to be an important characteristics of many coronaviruses. To account for this characteristic, we, assumed that secondary cases are generated according to a negative binomial distribution: 

$$I_t \sim NegBin\left( R_t \Sigma_{s=1}^{t} I_{t-s} w_s, z \right) $$

The value of the overdispersion parameter, z, was taken from analyses of exposure patterns during the SARS epidemic [lloyds-smith sars nature]. 
 
#### Projections 

Below for each country with active transmission (see methods), we plot the observed incidence of deaths in the last 3 weeks (black dots). Past forecasts, where available, are shown in green (median and 95%CrI band), while latest forecasts are show in red. Horizontal dashed line show the start and end of each week (Monday to Sunday).

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}
obs <- drake::readd("obs")
pred <- drake::readd("model_1")
pred <- pred[pred$si == "si_1", ]
pred$date <- as.Date(pred$date)
projection_plot(obs, pred)

```

#### Effective Reproduction Number Estimates

In the figure below, for each country with active transmission (see methods), we plot the latest estimate of transmissibility ($R_t$), specifically estimated median and associated 95%CrI.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}
rt <- drake::readd("model_1_rt")
rt <- rt[rt$si == "si_1", ]
rt <- rt[rt$proj == "RtI0_Std_results_week_end_2020-03-22", ]
rt_plot(rt)

```

#### Summary of results

Below for each country with active transmission (see methods), a table summarising for each period and country forecasted, the observed (where available) and forecasted weekly death count, and estimated level of transmissibility.

```{r echo = FALSE, warning = FALSE, message = FALSE}
x <- drake::readd("formatted_weekly_predictions_qntls")[[1]]
x <- x[grep("RtI0", x$model), ]
x <- dplyr::select(x, -model)
knitr::kable(
    x,
    ##align = "l",
    digits = 2,
    ##table.attr='class="table-fixed-header"'
    ) %>%
    kable_styling(
        bootstrap_options = "striped",
        ##full_width = F,
        ##position = "center",
        fixed_thead = TRUE
    ) %>% row_spec(0, align = "left")

```


### Model 2 {.tabset .tabset-fade .tabset-pills} 

#### Methods

**Estimating current transmissibility**

The methods is very similar to that of model but given uncertainty surrounding the epidemiological situation before the calibration period, we only used incidence data during the calibration period, and reconstructed the incidence before the calibration period whilst estimating R, as described in [Nouvellet 2018 simple approach]. The model has the advantage of being more robust to changes in reporting before the calibration.


**Forward projections**

As for model 1, we used the renewal equation to project the incidence forward, given a back-calculated early incidence curve, an estimated reproduction number, and the observed incidence over the calibration period. We sampled a sets of back-calculated early incidence curves and reproduction numbers from the posterior distribution obtained in the estimation process. For each of these sets, we simulated stochastic realisations of the renewal equation from the end of the calibration period; leading to projected incidence trajectories. 

Projections were made on a 7-days horizon. The projections assume that the transmissibility remains constant over this horizon. If transmissibility were to decrease as a result of control interventions and/or changes in behaviour over this time period, we would predict a lower number of cases; similarly, if transmissibility were to increase over this time period, we would predict a higher number of cases. We limited our projection to 7 days only as assuming constant transmissibility over longer time horizons seemed unrealistic. 



#### Projections

Below for each country with active transmission (see methods), we plot the observed incidence of deaths in the last 3 weeks (black dots). Past forecasts, where available, are shown in green (median and 95%CrI band), while latest forecasts are show in red. Horizontal dashed line show the start and end of each week (Monday to Sunday).


```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}
pred <- drake::readd("model_2") 
pred <- pred[pred$si == "si_1", ]
pred$date <- as.Date(pred$date)
projection_plot(obs, pred)
```

#### Effective Reproduction Number Estimates

In the figure below, for each country with active transmission (see methods), we plot the latest estimate of transmissibility ($R_t$), specifically estimated median and associated 95%CrI.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}

rt <- drake::readd("model_2_rt")
rt <- rt[rt$si == "si_1", ]
rt <- rt[rt$proj == "sbkp_Std_results_week_end_2020-03-22", ]
rt_plot(rt)

```

#### Summary of results

Below for each country with active transmission (see methods), a table summarising for each period and country forecasted, the observed (where available) and forecasted weekly death count, and estimated level of transmissibility.

```{r echo = FALSE, warning = FALSE, message = FALSE}
x <- drake::readd("formatted_weekly_predictions_qntls")[[1]]
x <- x[grep("sbkp", x$model), ]
x <- dplyr::select(x, -model)
knitr::kable(
    x,
    ##align = "l",
    digits = 2
    ) %>% kable_styling() %>%
     row_spec(0, align = "left")

```


## Sensitivity Analyses


Results with a longer Serial Interval (mean 6.48 days, SD 3.83 days).

### Projections

Below for each country with active transmission (see methods), we plot the observed incidence of deaths in the last 3 weeks (black dots). Past forecasts, where available, are shown in green (median and 95%CrI band), while latest forecasts are show in red. Horizontal dashed line show the start and end of each week (Monday to Sunday).

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}

ensb_pred2 <- ensb_pred[ensb_pred$si == "si_2", ]
ensb_pred2$date <- as.Date(ensb_pred2$date)
projection_plot(obs, ensb_pred2)

```

### Effective Reproduction Number Estimates

In the figure below, for each country with active transmission (see methods), we plot the latest estimate of transmissibility ($R_t$), specifically estimated median and associated 95%CrI.

## Case Ascertainment

The analysis rely on 2 key assumptions:
  
*	The true underlying Case Fatality Ratio (CFR) is 1%,
*	The delay from a death to this case being reported is 10 days.

First, we analyses the ratio of reported deaths to reported cases 10 days prior.
Therefore, the ratio roughly accounts for the delay between death and case being reported. 

Any temporal trend suggests a change in the reporting. 
For instance, an increase in the ratio gives an indication that cases reporting is decreasing.


**Temporal trend in the ratio of deaths to reported cases**

Sarting in march, we compute the average and 95%CI for the ratio of deaths  to reported cases (with a 10-days delay) using a moving window of 7 days.


```{r echo = FALSE, warning = FALSE, message = FALSE}
date_week_finishing <-  as.Date('22/03/2020',format = '%d/%m/%Y')
delay_report_death <- 10 # need checking!!

day.project <- 7
t.window.range <- 7

rep <- 2e4
d <- readRDS(file = here::here(paste0('Team.input/data_',date_week_finishing,'.rds')))

D <- d$D_active_transmission
I <- d$I_active_transmission
country <- d$Country
N_geo <- length(country)
date_week_finishing <- d$date_week_finishing

##colSums(D[,-1])/colSums(I[,-1])

date <- D$dates[7:(nrow(D)-delay_report_death)]
moving_DtoI <- list()

med <- matrix(NA,length(date),N_geo)
low <- med
up <- med

for (i in 1:nrow(med)){
  x = colSums(D[i:(i+6)+delay_report_death,-1])
  n = colSums(I[(i:(i+6)),-1])
  temp <- Hmisc::binconf(x = x , 
                  n = n,alpha = .05, method = 'exact')
  
  temp[which(x>n),] <- 1
  temp[which(n==0),] <- NA
  
  med[i,] <- temp[,1]
  low[i,] <- temp[,2]
  up[i,] <- temp[,3]
}

# med[which((med==Inf))] <- NA

for (i in 1:N_geo){
  moving_DtoI[[as.character(country[i])]] <- data.frame(date = date,
                                                        ratio = med[,i],
                                                        lower = low[,i],
                                                        upper = up[,i])
}

```

The figure below shows the temporal trends in the ratio of reported deaths to reported cases 10 days prior (medians and 95%CIs, solida lines and bands).

We also plot both the reported deaths (red) and reported cases (black). Those have
been rescaled but are comparable to each other. The rescaling is such
that the maximium recorded numbers of deaths or cases (with a 10-days
delay) reaches 1.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6.4}
f <- which(moving_DtoI[[1]]$date >= as.Date(c('01/03/2020'),format = '%d/%m/%Y'))

layout(matrix(1:4,2,2))
for (i in 1:N_geo){
  a <- moving_DtoI[[i]]
  plot(a$date,a$ratio,lwd = 2,
       #ylim = c(0,max(a[f,2:4],na.rm=TRUE)),
       ylim=c(0,1),
       type='l',
       xlim = c(as.Date(c('01/03/2020'),format = '%d/%m/%Y'),date_week_finishing-delay_report_death),
       bty ='n', main = country[i],col = rgb(0,0,1),
       xlab = '', ylab = 'ratio D to I') 
  
  polygon(c(a$date,rev(a$date)),
          c(a$lower,rev(a$upper)),
          border = NA,
          col = rgb(0,0,1,0.2))
  f2 <- which( I$dates %in% a$date)
  
  f2<-f2[f]
  inc <- cbind(I[f2,i+1],D[f2+delay_report_death,i+1])
  lines(a$date[f], 
        inc[,1]/max(c(inc[,1],inc[,2])),
        type = 'p', pch=16,col='black')
  
  lines(a$date[f], 
        inc[,2]/max(c(inc[,1],inc[,2])), 
        type = 'p', pch=16,col='red')
  
  if(i==3){
  legend('topleft',legend = c('ratio','death','reported cases'),bty='n',
         lwd=c(3,NA,NA),pch=c(NA,16,16),col = c(rgb(0,0,1),rgb(1,0,0),rgb(0,0,0)))
  }
}

```

*Note that if deaths exceed the number of reported cases 10-days before, we set the reporting at 1 (95%CI [1;1])*

**Estimating ascertainment**

If all cases (incl. asymptomatics) and deaths were reported, and the delay from reporting to death was exactly 10 days, then the ratio defined above would be equivalent to the Case Fatality Ratio (CFR).

Assuming a CFR of 1%, the ascertainment, $\rho$, can be defined as:
$$\rho = \frac{I^r}{D^r} CFR$$
with $D^r$ and $I^r$ the reported number of deaths and cases (with a 10-days delay).

And the ratio previously defined is 
$$\frac{D^r}{I^r}$$

Below we summarise:

* Estimated ratio per country (mean and 95%CI) during the last 14 days, 
* Estimated case ascertainment per country (mean and 95%CI) during the last 14 days,
* Estimated factor to be multipled to observed cases number to obatin true number of cases. Mean and 95%CI, estimated for the last 14 days.
* Observed and predicted number of cases in the last week (week starting [date]).

Average and 95%CI for ratio of deaths to reported cases for the last 14 days.
Suprisingly, this appear to be pretty much the case for South Korea
(see below).

```{r echo = FALSE, warning = FALSE, message = FALSE}
x <- readr::read_csv(here::here("Team.output/summary_DeathToRepoted_14days.csv"))
knitr::kable(x[, -1]) %>%
    kable_styling() 

```


## References

The forecasts produced relies on reported daily counts of deaths per country available on the eCDC website:

https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide

Other references:

* Cori A, Ferguson NM, Fraser C, Cauchemez S. A New Framework and Software to Estimate Time-Varying Reproduction Numbers During Epidemics. Am J Epidemiol. 2013; 178: 1505-1512.

* Wang Y, Teunis T. Strongly heterogeneous transmission of COVID-19 in mainland China: local and regional variation. March 10, 2020. medRxiv

* Ferguson et al. 2020. Report 9: Impact of non-pharmaceutical interventions (NPIs) to reduce COVID-19 mortality and healthcare demand. https://www.imperial.ac.uk/mrc-global-infectious-disease-analysis/news--wuhan-coronavirus/

* Lloyd-Smith  JO, Schreiber SJ, Kopp PE, Getz WM. Superspreading and the effect of individual variation on disease emergence. Nature, 2005; 438, 355-359.

* Nouvellet P et al. A simple approach to measure transmissibility and forecast incidence. Epidemics, 2018; 22. 29-35.
