---
output: 
pdf_document:
citation_package: natbib
keep_tex: false
fig_caption: true
latex_engine: pdflatex
title: "Short-term forecasts of cases and deaths of COVID-19 in multiple countries"
author:
- name: Sangeeta Bhatia and Pierre Nouvellet
affiliation: Imperial College London
abstract: 
keywords: 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: double
bibliography: 
biblio-style: apsr
endnote: no
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggpubr)
library(drake)
raw_data <- drake::readd(raw_data)
raw_data <- dplyr::rename(raw_data, country = "Countries.and.territories")


```
# Introduction {.tabset .tabset-fade .tabset-pills}

As of `r format(Sys.time(), '%B %d, %Y')`, there have more than
351,000 cases of COVID-19 across the world, with more than 15,000
deaths. In this report, we use the reported number of deaths
of COVID-19 to make short term forecasts of the trajectory of the
outbreak. 

## Methods 

The objective of this project is to forecast for the daily number of COVID-19 Death in countries with sustained local transmission. We aim to produce weekly forecasts every Monday. Multiple teams are involved, each producing weekly forecasts and a centralised team also produces an ensemble forecast.

Forecasts are produce weekly for based on data available on the last Sunday of the previous week. Forecasts are produced only for countries that reported at least 5 deaths in the previous week and 10 deaths in the previous two weeks. For the purpose of our analysis, we assume those countries show strong evidence active and sustained local transmission.

Importantly, all weekly forecasts rely on estimation of current level of transmissibility. Given we are forecasting number of daily deaths expected, current transmissibility reflects epidemiological dynamics when the cases who died became infected. 

Therefore, current estimates of transmissibility based on death reflect transmission events that occur about two to three weeks ago, given the delays between infection and death. As a consequence, in the best case scenario, the impact of current control on estimated of transmissibility will be quantifiable with a two to three weeks delay.

In the results section of each model output, a 'methods' section outlines briefly the methodology used to produce forecasts.


## Data Sources


The forecasts produced relies on reported daily counts of deaths per country available on the eCDC website:

https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide

While death counts are used for the main analysis and producing forecasts. Reported cases and the death to reported cases ratios for each country are show in the 'Supplementary Information' section.



## Results {.tabset .tabset-fade .tabset-pills} 

### Ensemble Model 

### Model 1 {.tabset .tabset-fade .tabset-pills} 

#### Cases - TO BE REMOVED

```{r echo = FALSE}
x <- drake::readd("model_predictions_qntls")[[1]]
x$date <- as.Date(x$date)
raw_data <- dplyr::filter(raw_data, country %in% x$country)

p <- ggplot() +
        geom_point(data = raw_data, aes(DateRep, Cases)) +
        geom_line(data = x, aes(date, `50%`, group = country)) +
    geom_ribbon(data = x,
                aes(x = date,
                    ymin = `2.5%`,
                    ymax = `97.5%`,
                    group = country),
                alpha = 0.3
                ) +
        geom_ribbon(data = x,
                aes(x = date,
                    ymin = `40%`,
                    ymax = `60%`,
                    group = country),
                alpha = 0.4
                ) +
    facet_wrap(~country, scales = "free_y") +
    theme_pubr() +
    xlab("") + ylab("Cases")

p
    
```

#### Deaths

#### Effective Reproduction Number Estimates

### Summary of results

a large table with a row per country that ever appeared as 'active transmission'
and the following columns:
- country name

for each retrospective week of predictions

- Rt estimated on the first week of predictions (week starting 9th of march) - paste0 with median, 95%CrI(XX;XX)
- predicted weekly death count - median, 95%CrI(XX;XX)
- observed weekly death count

for the last week of predictions:
- Rt estimated on the first week of predictions (week starting 9th of march) - paste0 with median, 95%CrI(XX;XX)
- predicted weekly death count - median, 95%CrI(XX;XX)

ideally downloadable as xlsx file formatted to keep country column and title row fixed

### Model 2 

"Sed ut perspiciatis unde omnis iste natus error sit voluptatem
accusantium doloremque laudantium, totam rem aperiam, eaque ipsa quae
ab illo inventore veritatis et quasi architecto beatae vitae dicta
sunt explicabo. Nemo enim ipsam voluptatem quia voluptas sit
aspernatur aut odit aut fugit, sed quia consequuntur magni dolores eos
qui ratione voluptatem sequi nesciunt. Neque porro quisquam est, qui
dolorem ipsum quia dolor sit amet, consectetur, adipisci velit, sed
quia non numquam eius modi tempora incidunt ut labore et dolore magnam
aliquam quaerat voluptatem. Ut enim ad minima veniam, quis nostrum
exercitationem ullam corporis suscipit laboriosam, nisi ut aliquid ex
ea commodi consequatur? Quis autem vel eum iure reprehenderit qui in
ea voluptate velit esse quam nihil molestiae consequatur, vel illum
qui dolorem eum fugiat quo voluptas nulla pariatur?"

### Model 3 

"Sed ut perspiciatis unde omnis iste natus error sit voluptatem
accusantium doloremque laudantium, totam rem aperiam, eaque ipsa quae
ab illo inventore veritatis et quasi architecto beatae vitae dicta
sunt explicabo. Nemo enim ipsam voluptatem quia voluptas sit
aspernatur aut odit aut fugit, sed quia consequuntur magni dolores eos
qui ratione voluptatem sequi nesciunt. Neque porro quisquam est, qui
dolorem ipsum quia dolor sit amet, consectetur, adipisci velit, sed
quia non numquam eius modi tempora incidunt ut labore et dolore magnam
aliquam quaerat voluptatem. Ut enim ad minima veniam, quis nostrum
exercitationem ullam corporis suscipit laboriosam, nisi ut aliquid ex
ea commodi consequatur? Quis autem vel eum iure reprehenderit qui in
ea voluptate velit esse quam nihil molestiae consequatur, vel illum
qui dolorem eum fugiat quo voluptas nulla pariatur?"


## Supplementary Information

To include graph of reported cases

To include analysis and graph of death to reported cases ratio