---
title: "ccc"
author: "Pierre Nouvellet"
date: "2019-07"
---

```{r options, include = FALSE, message = FALSE, warning = FALSE, error = FALSE}
library(knitr)
library(Hmisc)

opts_chunk$set(collapse = TRUE)

opts_chunk$set(fig.path='figs/', fig.keep='high', 
               dev=c('png'), fig.width=10, fig.height=6, cache=FALSE,
               tidy=FALSE, warning=FALSE, fig.show="asis"
 	       )

```


NEED to do:
1) input

# Input

```{r}

date_week_finishing <-  as.Date('22/03/2020',format = '%d/%m/%Y')
delay_report_death <- 10 # need checking!!

day.project <- 7
t.window.range <- 7

rep <- 2e4

```

# Read data

```{r}
d <- readRDS(file = paste0('../Team.input/data_',date_week_finishing,'.rds'))

D <- d$D_active_transmission
I <- d$I_active_transmission
country <- d$Country
N_geo <- length(country)
date_week_finishing <- d$date_week_finishing

```


```{r}

colSums(D[,-1])/colSums(I[,-1])

date <- D$dates[7:(nrow(D)-delay_report_death)]
moving_DtoI <- list()

med <- matrix(NA,length(date),N_geo)
low <- med
up <- med

for (i in 1:nrow(med)){
  x = colSums(D[i:(i+6)+delay_report_death,-1])
  n = colSums(I[(i:(i+6)),-1])
  temp <- binconf(x = x , 
                  n = n,alpha = .05, method = 'exact')
  
  temp[which(x>n),] <- 1
  temp[which(n==0),] <- NA
  
  med[i,] <- temp[,1]
  low[i,] <- temp[,2]
  up[i,] <- temp[,3]
}

# med[which((med==Inf))] <- NA

for (i in 1:N_geo){
  moving_DtoI[[as.character(country[i])]] <- data.frame(date = date,
                                                        ratio = med[,i],
                                                        lower = low[,i],
                                                        upper = up[,i])
}

```

Average and 95%CI for ratio of deaths (with a 10-days delay) to reported cases for moving window of 7 days since march.

The ratio roughly accounts for the delay between death and case being reported. 



When deaths exceed the number of reported cases 10-days before, we set the reporting at 1 (95%CI [1;1])



Any temporal trend suggests a change in the reporting. 
For instance, an increase in the ratio gives an indication that cases reporting is decreasing.


In the plots below, both reported deaths (red) and cases (black) have been rescaled but are comparable to each other. The rescaling is such that the maximium recorded numbers of death or cases (with a 10-days delay) reaches 1.

```{r}

f <- which(moving_DtoI[[1]]$date >= as.Date(c('01/03/2020'),format = '%d/%m/%Y'))

layout(matrix(1:4,2,2))
for (i in 1:N_geo){
  a <- moving_DtoI[[i]]
  plot(a$date,a$ratio,lwd = 2,
       #ylim = c(0,max(a[f,2:4],na.rm=TRUE)),
       ylim=c(0,1),
       type='l',
       xlim = c(as.Date(c('01/03/2020'),format = '%d/%m/%Y'),date_week_finishing-delay_report_death),
       bty ='n', main = country[i],col = rgb(0,0,1),
       xlab = '', ylab = 'ratio D to I') 
  
  polygon(c(a$date,rev(a$date)),
          c(a$lower,rev(a$upper)),
          border = NA,
          col = rgb(0,0,1,0.2))
  f2 <- which( I$dates %in% a$date)
  
  f2<-f2[f]
  inc <- cbind(I[f2,i+1],D[f2+delay_report_death,i+1])
  lines(a$date[f], 
        inc[,1]/max(c(inc[,1],inc[,2])),
        type = 'p', pch=16,col='black')
  
  lines(a$date[f], 
        inc[,2]/max(c(inc[,1],inc[,2])), 
        type = 'p', pch=16,col='red')
  
  if(i==3){
  legend('topleft',legend = c('ratio','death','reported cases'),bty='n',
         lwd=c(3,NA,NA),pch=c(NA,16,16),col = c(rgb(0,0,1),rgb(1,0,0),rgb(0,0,0)))
  }
}

```

Average and 95%CI for ratio of deaths to reported cases for the last 14 days.


If all cases (incl. asymptomatics) and death were reported, and the delay fro reporting was exactly 10 days, then the ratio defined would be equivalent to the CFR.

Suprisingly, this appear to be pretty much the case for South Korea (see below).

```{r}

limits <- seq(date_week_finishing-13,date_week_finishing,by = 1)
f <- which(D$dates %in% limits)
temp <- binconf(x = colSums(D[f,-1]) , n = colSums(I[f-delay_report_death,-1]),alpha = .05, method = 'exact')


# assumed CFR
CFR <- .01
rho <- 1/temp*CFR
rho[rho>1]<-1

# meaning the real epidemics is x times bigger than observed
multi_real <- 1/rho

# summary
summary_14days <- data.frame(country = country,
                      death = colSums(D[f,-1]),
                      cases = colSums(I[f,-1]),
                      deaths_to_reported_ratio = paste0(round(temp[,1],digits = 3)*100,
                                                        '% ; 95%CI [',
                                                        round(temp[,2],digits = 3)*100,
                                                        ';',
                                                        round(temp[,3],digits = 3)*100,
                                                        ']%'),
                      estimated_reporting = paste0(round(rho[,1],digits = 3)*100,
                                                        '% ; 95%CI [',
                                                        round(rho[,2],digits = 3)*100,
                                                        ';',
                                                        round(rho[,3],digits = 3)*100,
                                                        ']%'),
                      factor_to_real_size = paste0(round(multi_real[,1],digits = 1),
                                                        ' ; 95%CI [',
                                                        round(multi_real[,2],digits = 1),
                                                        ';',
                                                        round(multi_real[,3],digits = 1),
                                                        ']'))

summary_14days <- summary_14days[order(temp[,1],decreasing = TRUE),]

write.csv(summary_14days,file = 'summary_DeathToRepoted_14days.csv')

```

